QPLIB: A Library of Quadratic Programming Instances

### 2016-12-18 QPLIB-3.tex
Update QPLIB-3.tex
Finished checking all of §3; mostly cosmetic changes, with a few comments. AF

During the development of the library, a discussion ensued about what
the expected gols that we wanted to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item represent as much as possible all the different categories of QP
       problems;
 %
 \item gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item include for each of the categories a set of well-diversified
       instances;
 %
 \item obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve the aforementioned goals, we performed the following two
filters, applied in cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S \ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in {\tt Gams} (cf.~Table \ref{t:solvers})
       capable of solving it, whose number depends on the category of the
       instance under consideration. We then filtered according to a relative
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives. Finally, in order to only keep computationally
       challenging instances we ran a general purpose solver able to tackle
       all QP categories with a time limit of 120 seconds; all the instances
       which have been solved to proven optimality within this time limit were
       discarded.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

@@

In this section we analyse the feature of the instances selected to be
part of library.
The characteristics of the instances in the final library are
presented in Table \ref{tab:DD} for \emph{discrete} instances
(*\{B,M,I,G\}*) and in Table \ref{tab:CC} for
\emph{continuous} ones (*C*).
For each category, the Tables report in column $\#$ the corresponding
number of instances.
The final set well respects the original distribution of the gathered
instances among the different categories.
It is worth noticing that the classical discrete categories (such as
(L,M,Q) or (Q,B,L)) are well represented by $118$ and $59$  instances
respectively. On the other side, the classical continuous categories
(such as (L,C,Q) and (QCQ)) are also well represented by $50$ and $17$
 instances respectively.
Moreover, the library covers a large majority of all possible
categories of instances.

\begin{center}
\begin{table}
 \centering
 %\scriptsize
 \setlength{\tabcolsep}{18pt}
 \renewcommand \arraystretch{1.1}
\begin{tabular}{lllr}
\toprule
Obj. Fun. & Variables & Constraints & \#\\
\cmidrule(lr){1-4}
%
\multirow{5}*{Linear}
          & \multirow{1}*{Binary}
%                    & None      &   \\
%          &         & Linear    &  \\
                    & Quadratic &   9 \\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{2}*{Mixed}
                    & Convex    &   2\\[1.2 ex]
          &         & Quadratic &    118\\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{1}*{Integer}
%                    & Linear    &    \\
                   & Quadratic &    2\\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{1}*{General}
%                    & Linear    &    \\
                   & Quadratic &    3\\[1.2 ex]
\cmidrule(lr){1-4}
%\multirow{5}*{Linear}
%          & Binary  & Quadratic &   9\\
%          & \multirow{2}*{Mixed}
%                    & Convex    &   2\\
%          &         & Quadratic &  118\\
%%          & \multirow{2}*{Integer}
%%                    & Convex    &  15 \\
%%          &         & Quadratic &   5 \\
%          & {General} & Quadratic &   3 \\
%\hline
\multirow{3}*{Convex}
          & Binary  & Linear    &  2 \\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{2}*{Mixed}
                    & Linear    &   13\\[1.2 ex]
          &         & Quadratic &    1\\[1.2 ex]
%          & General & Linear    &    \\
%\hline
\cmidrule(lr){1-4}
\multirow{7}*{Quadratic}
          & \multirow{3}*{Binary}
                    & None      &   23\\[1.2 ex]
          &         & Linear    &  59\\[1.2 ex]
          &         & Quadratic &   5 \\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{2}*{Mixed}
                    & Linear    &   10\\[1.2 ex]
          &         & Quadratic &    1\\[1.2 ex]
%          & Integer & Linear    &    \\
\cmidrule(lr){2-4}
          & \multirow{1}*{Integer}
                    & Linear    &    2\\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{1}*{General}
                    & Quadratic    &    1\\[1.2 ex]
%          &         & Quadratic &    \\
\hline
Total     &         &           & 251\\
%
\bottomrule
\end{tabular}
\caption{Classification of the final set of discrete instances}\label{tab:DD}
\end{table}
\end{center}

\begin{center}
\begin{table}
 \centering
 %\scriptsize
 \setlength{\tabcolsep}{18pt}
 \renewcommand \arraystretch{1.1}
\begin{tabular}{llr}
\toprule
Obj. Fun. & Constraints & \#\\
\cmidrule(lr){1-3}
%
\multirow{2}*{Linear}    & Convex    &   11\\[1.2 ex]
                         & Quadratic &   50\\[1.2 ex]
\cmidrule(lr){1-3}
\multirow{4}*{Convex}
                         & Box       &   3 \\[1.2 ex]
                         & Linear    &   12\\[1.2 ex]
                         & Convex    &    2\\[1.2 ex]
                         & Quadratic &    5\\[1.2 ex]
\cmidrule(lr){1-3}
\multirow{3}*{Quadratic}
                         & Linear    &   5\\[1.2 ex]
                         & Convex    &   11\\[1.2 ex]
                         & Quadratic &   17\\[1.2 ex]
\hline
Total                    &           & 116 \\
%
\bottomrule
\end{tabular}
\caption{Classification of the final set of continuous instances}\label{tab:CC}
\end{table}
\end{center}

To graphically represent the library we report some
graphical representations of the instance main features.

In Figure~\ref{fig:distribution} we represent 
on the horizontal (resp. vertical) axis the number of variables (resp.
constrains). Both axis are in logarithmic scale.
The ``$+$'' points represent  the continuous instances, while the 
``$\times$'' points represent the discrete instances. The figure shows that
the library contains instances of different size in terms of variables
and constraints. The number of constraints goes from few units up to
one hundred thousands constraints, while the number of variables goes
up to almost forty-five thousands.

In Figures~\ref{fig:pic_var_small}~\ref{fig:pic_var_medium}~\ref{fig:pic_var_large},
for each instances we report the total number of variables with a
``$+$'' point and  the total number of discrete variables with a
``$\times$'' point (i.e. binary and integer variables). The instances are
sorted accordingly to the total number of variables and, for reason of
readability, we splitted the figure into three sub figures. The first one hundred
instances are presented in Figures~\ref{fig:pic_var_small},
then the second two hundred instances are presented in
Figure~\ref{fig:pic_var_medium} and finally, the remaining $67$
instances are presented in Figure~\ref{fig:pic_var_large}.
From the picture we evince that for all instance dimensions it is
possible to find discrete instances with different percentages of
continuous variables and purely continuous ones.

In Figures~\ref{fig:pic_constr_small}~\ref{fig:pic_constr_medium}~\ref{fig:pic_constr_big},
for each instances we report the total number of constraints with a
``$+$'' point and the total number of quadratic (either convex of non
convex) constraints with a ``$\times$'' point. The instances are sorted
accordingly to the total number of constraints  and divided in the
same manner as for Figures~\ref{fig:pic_var_small}~\ref{fig:pic_var_medium}~\ref{fig:pic_var_large}.
Also in this case, we can see that for all sizes it is possible to
find instances with only linear constraints and instances with
different percentage of linear and quadratic constraints.

In Figures~\ref{fig:pic_density} and~\ref{fig:pic_neg_eig},  we
focused on the instances with a quadratic objective function.
In Figures~\ref{fig:pic_density} we plot the density of the $Q^0$
matrix and we sort the instance according to it. We observe that the
majority of the instances have either a low or high density. However,
also intermediate values are present. It is worth noticing that around 60 instances have a 
density less that 5\% (see the Appendix for further details). 
In Figures~\ref{fig:pic_neg_eig} we present  the percentage of
negative eigenvalues and we sort the instance according to it. As
expected, a vast majority of the instances have $50\% $ of negative
eigenvalues. In addition, around $40$ instances are convex (i.e. $0\%$
of negative eigenvalues) but also instances of high or low percentages
of negative eigenvalues are present in the library.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

@@

@@

@@

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_constr_big.png}
  \caption{Number of constraints (c) \label{fig:pic_constr_big}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{figure}\centering
%  \includegraphics[width=0.85\textwidth]{pic_quad_vs_convex_constr.png}
%  \caption{Number of quadratic constraints (convex and non-convex)
%\label{fig:pic_quad_vs_convex_constr}}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{figure}\centering
%  \includegraphics[width=0.85\textwidth]{pic_fuffa.png}
%  \caption{...\label{fig:9}}
%\end{figure}

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

>>>>

During the development of the library, a discussion ensued about what
the expected gols that we wanted to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item represent as much as possible all the different categories of QP
       problems;
 %
 \item gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item include for each of the categories a set of well-diversified
       instances;
 %
 \item obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve the aforementioned goals, we performed the following two
filters, applied in cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S \ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in {\tt Gams} (cf.~Table \ref{t:solvers})
       capable of solving it, whose number depends on the category of the
       instance under consideration. We then filtered according to a relative
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives. {\bf Comment: we may want to be a tad more specific
       about what ``features'' exactly meant there}. Finally, in order to only
       keep computationally challenging instances we ran a general purpose solver
       able to tackle all QP categories with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded. {\bf Comment: this might actually be best characterized as
       a third filter.}
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

@@

We now analyse the features of the instances selected to be part of the
library. The characteristics of the instances are presented in Table
\ref{tab:DD} for \emph{discrete} instances (*\{B,M,I,G\}*) and in Table
\ref{tab:CC} for \emph{continuous} ones (*C*). For each category, the Tables
report in column $\#$ the corresponding number of instances. It can be seen
that the final set well respects the original distribution of the gathered
instances among the different categories. Indeed, the discrete categories
(LMQ) or (QBL) are well represented by $118$ and $59$ instances,
respectively. Similarly, the continuous categories (LCQ) and (QCQ) are well
represented by $50$ and $17$ instances, respectively. Moreover, the library
actually covers the large majority of all possible categories of instances.

\begin{center}
\begin{table}
 \centering
 %\scriptsize
 \setlength{\tabcolsep}{18pt}
 \renewcommand \arraystretch{1.1}
\begin{tabular}{lllr}
\toprule
Obj. Fun. & Variables & Constraints & \#\\
\cmidrule(lr){1-4}
%
\multirow{5}*{Linear}
          & \multirow{1}*{Binary}
%                    & None      &   \\
%          &         & Linear    &  \\
                    & Quadratic &   9 \\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{2}*{Mixed}
                    & Convex    &   2\\[1.2 ex]
          &         & Quadratic &    118\\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{1}*{Integer}
%                    & Linear    &    \\
                   & Quadratic &    2\\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{1}*{General}
%                    & Linear    &    \\
                   & Quadratic &    3\\[1.2 ex]
\cmidrule(lr){1-4}
%\multirow{5}*{Linear}
%          & Binary  & Quadratic &   9\\
%          & \multirow{2}*{Mixed}
%                    & Convex    &   2\\
%          &         & Quadratic &  118\\
%%          & \multirow{2}*{Integer}
%%                    & Convex    &  15 \\
%%          &         & Quadratic &   5 \\
%          & {General} & Quadratic &   3 \\
%\hline
\multirow{3}*{Convex}
          & Binary  & Linear    &  2 \\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{2}*{Mixed}
                    & Linear    &   13\\[1.2 ex]
          &         & Quadratic &    1\\[1.2 ex]
%          & General & Linear    &    \\
%\hline
\cmidrule(lr){1-4}
\multirow{7}*{Quadratic}
          & \multirow{3}*{Binary}
                    & None      &   23\\[1.2 ex]
          &         & Linear    &  59\\[1.2 ex]
          &         & Quadratic &   5 \\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{2}*{Mixed}
                    & Linear    &   10\\[1.2 ex]
          &         & Quadratic &    1\\[1.2 ex]
%          & Integer & Linear    &    \\
\cmidrule(lr){2-4}
          & \multirow{1}*{Integer}
                    & Linear    &    2\\[1.2 ex]
\cmidrule(lr){2-4}
          & \multirow{1}*{General}
                    & Quadratic    &    1\\[1.2 ex]
%          &         & Quadratic &    \\
\hline
Total     &         &           & 251\\
%
\bottomrule
\end{tabular}
\caption{Classification of the final set of discrete instances}
\label{tab:DD}
\end{table}
\end{center}

\begin{center}
\begin{table}
 \centering
 %\scriptsize
 \setlength{\tabcolsep}{18pt}
 \renewcommand \arraystretch{1.1}
\begin{tabular}{llr}
\toprule
Obj. Fun. & Constraints & \#\\
\cmidrule(lr){1-3}
%
\multirow{2}*{Linear}    & Convex    &   11\\[1.2 ex]
                         & Quadratic &   50\\[1.2 ex]
\cmidrule(lr){1-3}
\multirow{4}*{Convex}
                         & Box       &   3 \\[1.2 ex]
                         & Linear    &   12\\[1.2 ex]
                         & Convex    &    2\\[1.2 ex]
                         & Quadratic &    5\\[1.2 ex]
\cmidrule(lr){1-3}
\multirow{3}*{Quadratic}
                         & Linear    &   5\\[1.2 ex]
                         & Convex    &   11\\[1.2 ex]
                         & Quadratic &   17\\[1.2 ex]
\hline
Total                    &           & 116 \\
%
\bottomrule
\end{tabular}
\caption{Classification of the final set of continuous instances}
\label{tab:CC}
\end{table}
\end{center}

We now report some graphs that help in illustrating the main features
of the instences. In Figure~\ref{fig:distribution} we plot the number
of variables (horizontal axis) versus the number of constrains
(vertical axis), both in logarithmic scale. Continuous instances are
marked with ``$+$'', and discrete ones with ``$\times$''. The figure shows
that the library contains a quite diverse set of instances of different in
terms of number of variables and constraints. The number of constraints goes
from few units up to one hundred thousands constraints, while the number of
variables goes up to almost forty-five thousands.

@@

Figure~\ref{fig:pic_var_small}, Figure~\ref{fig:pic_var_medium} and
Figure~\ref{fig:pic_var_large} describe how discrete and continuous
variables are distributed within the instances. The instances are
sorted accordingly to the total number of variables and, for reason of
readability, the graph is splitted in the three figures:
Figure~\ref{fig:pic_var_small} describes the 100 smaller instances,
Figure~\ref{fig:pic_var_medium} the 200 medium ones, and
Figure~\ref{fig:pic_var_large} the 67 largest ones. For each instance
we report the total number of variables with a ``$+$'', and the total
number of discrete variables (binary or general integer) with a
``$\times$''. The pictures clearly show that instances with different
percentages of integer and continuous variables are present in the
library, and well distributed across the whole spectrum of variable
sizes.

@@

Similarly, Figures~\ref{fig:pic_constr_small}, Figure~\ref{fig:pic_constr_medium}
and Figure~\ref{fig:pic_constr_big} describe how linear and quadratic
constraints are distributed within the instances. The instances are sorted
accordingly to the total number of constraints and divided in the
same manner as for Figures~\ref{fig:pic_var_small}--\ref{fig:pic_var_large}.
For each instances we report the total number of constraints with a ``$+$''
and the total number of (either convex of nonconvex) quadratic constraints
with a ``$\times$''. Also in this case, different percentages of linear and
quadratic constraints are present and well-distributed across the spectrum
of constraint sizes, although both medium- and large-size instances show a
prevalence of lower percentages of quadratic constraints. In particular,
from Figure~\ref{fig:pic_constr_big} we learn that while the maximum number
of linear constraints exceeds 100000, the maximum number of quadratic
constraints tops up at around 10000. This is, however, reasonable, considering
how quadratic constraints can, in general, be expected to be much more
computationally challenging than linear ones, especially if nonconvex.

@@

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_constr_big.png}
  \caption{Number of constraints (c) \label{fig:pic_constr_big}}
\end{figure}

Speaking of nonconvexity, Figure~\ref{fig:pic_neg_eig} focuses on
the instances with a quadratic objective function, ordered by the
the percentage of negative eigenvalues (vertical axis). Although
almost all percentages are represented, the instances are mostly
clustered around two values. About 25\% of the instances are convex,
i.e., they have 0\% of negative eigenvalues. Among the others, a vast
majority have around 50\% of negative eigenvalues. However, instances
with high or low percentages of negative eigenvalues are present.

Similarly, Figure~\ref{fig:pic_density} shows the instances with a
quadratic objective function sorted according to the density of the
$Q^0$ matrix (vertical axis). The majority of the instances have either
a very low or a rather high density: indeed, about 30\% of the instances
have density smaller that 5\%, and about 30\% of the instances have
density larger that 50\%. However, also intermediate values are present.

@@

{\bf Nothing about convexity of constraints? A graph plotting the
number of convex quadratic constraints vs the total number of quadratic
constraints?}
